Higher-Order unification for free!

### 2024-04-12 paper/main.tex
more work on abstract

\begin{abstract}
  Specifying and implementing a proof system from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level Meta Languages (ML) to facilitate this task in two
  key ways: 1) variable binding and substitution are simplified when ML binders
  represent object logic ones; 2) proof construction, and even proof search, is
  greatly simplified by leveraging the unification procedure provided by the ML.
  Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been utilized to implement various formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873},
  Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
  Constructions~\cite{felty93lics}.

  The object logic we are interested in is Coq's~\cite{Coq-refman}
  Calculus of Inductive Constructions (CIC),
  for which we aim to implement a unification procedure \Uo using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory comprises
  $\eta\beta$ equivalence and comes equipped with a
  higher order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  We want \Uo to be as powerful as \Ue but on the object logic CIC.
  Elpi also comes with an encoding for CIC that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo,
  ``underuses'' \Ue by restricting it to first-order unification problems only.
  To address this issue, we propose a better-behaved encoding, \Ho,
  demonstrate how to map unification problems in \Fo{}
  to related problems in \Ho, and illustrate
  how to map back the unifiers found by \Ue, effectively implementing
   \Uo on top of \Ue for the encoding \Fo.

  % We apply this technique to the implementation of a type-class~\cite{wadler89}
  % solver for Coq~\cite{Coq-refman}.
  % Type-class solvers are proof search procedures based on
  % unification that back-chain designated lemmas, providing essential
  % automation to widely used
  % Coq libraries such as Stdpp/Iris~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
  % and TLC~\cite{10.1007/978-3-642-14052-5_15}. These two libraries
  % constitute our test bed.
\end{abstract}

>>>>

\begin{abstract}
  Specifying and implementing a proof system from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level Meta Languages (ML) to facilitate this task in two
  key ways: 1) variable binding and substitution are for free when ML binders
  represent object logic ones; 2) proof construction, and even proof search, is
  greatly simplified by leveraging the unification procedure provided by the ML.
  Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been utilized to implement various formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873},
  Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
  Constructions~\cite{felty93lics}.

  The object logic we are interested in is % an extension of the latter,
  Coq's~\cite{Coq-refman} Calculus of Inductive Constructions (CIC), for which 
  we want to implement a unification based 
  proof search procedure using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory comprises
  $\eta\beta$ equivalence and comes equipped with a
  higher order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  Elpi also comes with an encoding for CIC that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo,
  ``underuses'' \Ue by restricting it to first-order unification problems only.

  In this paper we propose a better-behaved encoding, \Ho,
  and show how to map unification problems in \Fo{}
  to related problems in \Ho.
    As a result we can lift the ML
  higher-order unification procedure to \Fo{} in a way that it
  honours $\eta\beta$-equivalence (for CIC functions), solves
  problems in the pattern fragment and finally allows
  for the addition of heuristics to deal with problems outside the
  pattern fragment. The unification procedure we build can be used to
  effectively simulate a logic program in \Fo{} by re-using optimizations
  provided by the ML such as clause indexing.

\end{abstract}
